---

Part 2: Modeling and Evaluation
output:
  pdf_document: default
  html_document:
    code_folding: none
    theme: default
  html_notebook:
    code_folding: none
    theme: default
---


# Objective

This project is to build a linear regression model that can predict the `Gross` revenue earned by a movie based on other variables. Use R packages to fit and evaluate a regression model.



# Setup

load the dataset into memory:

```{r, echo=FALSE, warning=FALSE}
load('movies_merged')
```

This creates an object of the same name (`movies_merged`). For convenience, we can copy it to `df` and start using it:

```{r}
df = movies_merged
cat("Dataset has", dim(df)[1], "rows and", dim(df)[2], "columns", end="\n", file="")
colnames(df)
```

## Load R packages

Load  R packages that will need to use. 

```{r}
#install.packages("tidyr")

library(ggplot2)
library(stringr)
library(tidyr)
```



# Data Preprocessing

Beforestart building models, we should clean up the dataset and perform any preprocessing steps that may be necessary.
## 1. Remove non-movie rows

```{r}
#  Remove all rows from df that do not correspond to movies
df2 <- df[df$Type == "movie",]
```

## 2. Drop rows with missing `Gross` value

Since our goal is to model `Gross` revenue against other variables, rows that have missing `Gross` values are not useful to us.

```{r}
#  Remove rows with missing Gross value
df2<-df2[!is.na(df2$Gross),]
```

## 3. Exclude movies released prior to 2000

Inflation and other global financial factors may affect the revenue earned by movies during certain periods of time. Taking that into account is out of scope for this project, so let's exclude all movies that were released prior to the year 2000 ( use `Released`, `Date` or `Year` for this purpose).

```{r}
# Exclude movies released prior to 2000


rmlist<-NULL
totaln<- length(df2$Year)
for (n in 1:totaln){
   if ((as.numeric(df2$Year[n]))<2000)
   rmlist<-c(rmlist,n)
}
if (!is.null(rmlist))
  df2<-df2[-c(rmlist),]

```

## 4. Eliminate mismatched rows

_Note: You may compare the `Released` column (string representation of release date) with either `Year` or `Date` (numeric representation of the year) to find mismatches. The goal is to avoid removing more than 10% of the rows._

```{r}
# Remove mismatched rows
rmlist<-NULL
for (i in 1:length(df2$Title)){
  #step1. parsed Year, Date and Relaesed columns and get numeric value of years
  year <- as.numeric(df2$Year[i])
  date <- as.numeric(df2$Date[i])
  release <- as.numeric(format(df2$Released[i], format="%Y"))
 #step2. for each movie, need to check 3 pairs of year values for consistency:
  #(Date,Year)
  #(Date,Released)
  #(Year, Released)
  #if there is any "na"" value in a particular pair, then assume that pair is consistent;if the difference in the pair is greater than +/- one year,  then assume that pair is inconsistent; if movie has at least one inconsistent pairs, then that movie shall be removed

  if (!(is.na(year) ||  is.na(release))){
    if (( (year+1) < release ) || ((year-1) > release) )
      rmlist<-c(rmlist,i)}
  else if (!(is.na(date) ||  is.na(release))){
    if (( (date+1) < release ) || ((date-1) > release) )
      rmlist<-c(rmlist,i)}
  else if (!(is.na(date) ||  is.na(year))){
    if (( (date+1) < year ) || ((date-1) > year) )
      rmlist<-c(rmlist,i)}

}
#step3. remove rows in dataframe according to "rmlist":
rmlist <- unique(rmlist )
if (!is.null(rmlist))
  df2<-df2[-c(rmlist),]
```

## 5. Drop `Domestic_Gross` column

`Domestic_Gross` is basically the amount of revenue a movie earned within the US.I t is very highly correlated with `Gross` and is in fact equal to it for movies that were not released globally. Hence, it should be removed for modeling purposes.

```{r}
#  Exclude the `Domestic_Gross` column
df2<-df2[,-37]
```

## 6. Process `Runtime` column

```{r}
#  Replace df$Runtime with a numeric column containing the runtime in minutes

#first clear n/a values, if runtime is n/a , remove row:
nalist<-NULL
totaln<- length(df2$Runtime)
for (n in 1:totaln){
   if (df2$Runtime[n]=="N/A")
   nalist<-c(nalist,n)
}
if (!is.null(nalist))
  df2<-df2[-c(nalist),]
# now df2 does not contain n/a runtime

#next change runtime from string to numeric:
totaln<- length(df2$Runtime)
for (n in 1:totaln){
  
  nruntime <- unlist(strsplit(df2$Runtime[n], " "))
  
  if (is.element("h",nruntime) && is.element("min",nruntime)){
    nhour <- as.numeric(nruntime[1])
    nmin <- as.numeric(nruntime[3])
    totalmin <- nhour*60+nmin
  }  else if (is.element("h",nruntime) && (!is.element("min",nruntime))) {
    nhour <- as.numeric(nruntime[1])
    totalmin <- nhour*60
  } else {
    nmin <- as.numeric(nruntime[1])
    totalmin <- nmin
  }
  
  df2$Runtime[n]<- totalmin
  
}
df2$Runtime <- as.numeric(as.character(df2$Runtime))

```


```{r}
#  Additional preprocessing
#remove "na" imdb and tomato reviews and ratings

df2<-df2[!is.na(df2$imdbRating),]
df2<-df2[!is.na(df2$imdbVotes),]
df2<-df2[!is.na(df2$tomatoUserReviews),]
df2<-df2[!is.na(df2$tomatoUserRating),]
#below removes any movie with zero gross
rmlist<-NULL
for (i in 1:length(df2[,1]) ) {
  if (df2[i,37]==0)
  rmlist<- c(rmlist, i)
}
df2<-df2[-rmlist, ] 

```

## Final preprocessed dataset

Check the dimensions of the preprocessed dataset will be using for modeling and evaluation, and print all the final column names. 

```{r}
#  Print the dimensions of the final preprocessed dataset and column names
dim(df2)
names(df2)


```

# Cross Validation

In each of the tasks described in the next section, we will build a regression model. In order to compare their performance, we will compute the training and test Root Mean Squared Error (RMSE) at different training set sizes.

First, randomly sample 10-20% of the preprocessed dataset and keep that aside as the **test set**. The remainder of the preprocessed dataset is  **training data**.

## 1. Numeric variables

Use Linear Regression to predict `Gross` based on available _numeric_ variables. 

```{r}
#  Build & evaluate model 1 (numeric variables only)
################################################################
#first randomly select train dataset and test dataset 
selectrows<- function(data,percent){
  all<- seq.int(from = 1, to = length(data[,1])) 
  n<-as.integer(length(data[,1])*percent)
  select<-  sample(all, size = n)
  return(select)
}
set.seed(47)
trainrows<- selectrows(df2,0.9) # trainset uses 90% data, test set uses 10% data
trainset<- df2[trainrows,]
testset<- df2[-trainrows,]  

#################################################################################
#create a function to do training on the data
trainmodel1<- function(data){
    #model<- lm(Gross~Budget, data)
    #model<- lm(Gross~Budget+imdbRating, data)
    #model<- lm(Gross~Budget+imdbVotes, data)
    #model<- lm(Gross~Budget+imdbRating+imdbVotes+imdbRating*imdbVotes, data)
   model<- lm(Gross~Budget+Year+Runtime+imdbRating+imdbVotes+tomatoUserRating+tomatoUserReviews, data)
  return(model)
}
# models with j=20 various % of the data, repeat i=10 times for each %
trealgross<- testset[37]
trainrmse<- matrix( 1:40,  nrow=20, ncol=2) 
testrmse<- matrix( 1:40,  nrow=20, ncol=2) 
rmse<- 0
for (j in 1:20){ # use 20 different percentages
  lista<-NULL
  listb<-NULL
  for (i in 1:20){ # repeat 20 times
  select<-selectrows(trainset,j/20)
  data<- trainset[select,]
  M1<- trainmodel1(data)
  rmse<-sqrt(mean(residuals(M1)^2))
  lista<-c(lista,rmse)
  tpgross<- as.data.frame( predict(M1,testset))
  trmse<-0
  rss<- 0
  for (a in 1:length(testset[,1])){
  e<- (tpgross[a,1]-trealgross[a,1])^2
  rss<- rss+e
  }
  trmse<- sqrt(rss/length(testset[,1]))
  listb<-c(listb,trmse)
  }
  trainrmse[j,2]<-mean(lista) # this records all train rmse
  trainrmse[j,1]<- j*5 # this records the percentage of data used
  testrmse[j,2]<-mean(listb) # this records all train rmse
  testrmse[j,1]<- j*5 # this records the percentage of data used
}
testresult1<- as.data.frame(testrmse)
colnames(testresult1)<- c("percentage","RMSE")
trainresult1<- as.data.frame(trainrmse)
colnames(trainresult1)<- c("percentage","RMSE")
plotgraph<-function(testresult1,trainresult1){
ggplot(testresult1,aes(y=RMSE, x = percentage)) +  geom_point(data=testresult1,aes(y=RMSE, x = percentage, colour='test')) +  geom_line(data=testresult1,aes(y=RMSE, x = percentage, colour='test')) + geom_point(data=trainresult1,aes(y=RMSE, x = percentage, colour='train')) +  geom_line(data=trainresult1,aes(y=RMSE, x = percentage, colour='train'))
}

plotgraph(testresult1,trainresult1)

```



##  Feature transformations

Try to improve the prediction quality as much as possible by adding feature transformations of the numeric variables. Explore both numeric transformations such as power transforms and non-numeric transformations of the numeric variables like binning (e.g. `is_budget_greater_than_3M`).

```{r}
#  Build & evaluate model 2 (transformed numeric variables only)
#create a function to do training on the data
testset2<-testset
trainset2<-trainset

#bin budget into 4 category: 0 means lowbudget(<3M), 1 means midbudget(>3M but < 10M), 2 means highbudget (>10M but <100M) and 3 means superbudget(>100M) 
for (x in 1:length(testset2[,36])) {
  if (testset2[x,36]<=3000000)
    testset2[x,39]<- 0
  if (testset2[x,36]>3000000 && testset2[x,36]<=10000000)
    testset2[x,39]<- 1
  if (testset2[x,36]>10000000 && testset2[x,36]<=100000000)
    testset2[x,39]<- 2
  if (testset2[x,36]>100000000)
    testset2[x,39]<- 3
}

for (x in 1:length(trainset2[,36])) {
    if (trainset2[x,36]<=3000000)
    trainset2[x,39]<- 0
  if (trainset2[x,36]>3000000 && trainset2[x,36]<=10000000)
    trainset2[x,39]<- 1
  if (trainset2[x,36]>10000000 && trainset2[x,36]<=100000000)
    trainset2[x,39]<- 2
  if (trainset2[x,36]>100000000)
    trainset2[x,39]<- 3
}
colnames(testset2)[39]<- "budgetbin"
colnames(trainset2)[39]<- "budgetbin"

trainmodel2<- function(data){
    #model<- lm(Gross~Budget, data)
    #model<- lm(Gross~Budget+imdbRating, data)
    #model<- lm(Gross~Budget+imdbVotes, data)
    #model<- lm(Gross~Budget+imdbRating+imdbVotes+imdbRating*imdbVotes, data)
   #model<- lm(Gross~Budget+log(Budget)+Year+Runtime+log(Runtime)+imdbRating+imdbVotes+tomatoUserRating+tomatoUserReviews, data)
   model<- lm(Gross~Budget+I(Budget^2)+I(Budget^3)+Year+Runtime+I(log(Runtime))+imdbRating*imdbVotes+tomatoUserRating*tomatoUserReviews+budgetbin, data)
  return(model)
}
# models with j=20 various % of the data, repeat i=10 times for each %
trealgross<- testset[37]
trainrmse<- matrix( 1:40,  nrow=20, ncol=2) 
testrmse<- matrix( 1:40,  nrow=20, ncol=2) 
rmse<- 0
for (j in 1:20){ # use 19 different percentages
  lista<-NULL
  listb<-NULL
  for (i in 1:20){ # repeat 20 times
  select<-selectrows(trainset2,j/20)
  data<- trainset2[select,]
  M2<- trainmodel2(data)
  rmse<-sqrt(mean(residuals(M2)^2))
  lista<-c(lista,rmse)
  tpgross<- as.data.frame( predict(M2,testset2))
  trmse<-0
  rss<- 0
  for (a in 1:length(testset2[,1])){
  e<- (tpgross[a,1]-trealgross[a,1])^2
  rss<- rss+e
  }
  trmse<- sqrt(rss/length(testset2[,1]))
  listb<-c(listb,trmse)
  }
  trainrmse[j,2]<-mean(lista) # this records all train rmse
  trainrmse[j,1]<- j*5 # this records the percentage of data used
  testrmse[j,2]<-mean(listb) # this records all train rmse
  testrmse[j,1]<- j*5 # this records the percentage of data used
}
testresult2<- as.data.frame(testrmse)
colnames(testresult2)<- c("percentage","RMSE")
trainresult2<- as.data.frame(trainrmse)
colnames(trainresult2)<- c("percentage","RMSE")

plotgraph(testresult2,trainresult2)

###########################################################################

```

##  Non-numeric variables

Here we explore nin-numeric values, write code that converts genre, actors, directors, and other categorical variables to columns that can be used for regression . Also process variables such as awards into more useful columns (again, like you did in Project 1). Now use these converted columns only to build your next model.

```{r, echo=FALSE, warning=FALSE}
#  converted non-numeric variables only
#backup and save separate test and train data as below:
testset3<-testset2
trainset3<-trainset2
df3<-df2
#################################################################################
#split award into wins and nominations
processaward<- function(df){
wins <- as.numeric(str_extract(str_extract(df$Awards, '\\d+ wins?'), '\\d+'))
wins[is.na(wins)] <- 0
nominations <- as.numeric(str_extract(str_extract(df$Awards, '\\d+ nominations?'), '\\d+'))
nominations[is.na(nominations)] <- 0
won <- as.numeric(str_extract(str_extract(df$Awards, 'Won \\d+'), '\\d+'))
won[is.na(won)] <- 0
nominated <- as.numeric(str_extract(str_extract(df$Awards, 'Nominated for \\d+'), '\\d+'))
nominated[is.na(nominated)] <- 0
na <- df$Awards == 'N/A'
df$Wins <- wins + won
df$Wins[na] <- 0 #na become 0
df$Nominations <- nominations + nominated
df$Nominations[na] <- 0 #na become 0
return(df)
}

testset3<- processaward(testset3)
trainset3<- processaward(trainset3)
#################################################################################
#split genre
process<- function(dfa,dfb){
glist <- NULL
totaln <- length(dfa$Genre)
for (n in 1:totaln){
    gen <- unlist(strsplit(dfa$Genre[n], ", "))
    glist<- (c(glist,gen))
}

glist<-names(sort(summary(as.factor(glist)), decreasing=T)[1:11])
glist<-glist[-1]


# for every genre in glist, create a new colum in dataframe:
for (g in glist){
  colname <- g
  genrevalue<- NULL
  for (i in 1:length(dfb$Genre)){
    if (grepl(g, dfb$Genre[i], fixed=TRUE))
      genrevalue <- c(genrevalue,1)
    else genrevalue <- c(genrevalue,0)
  }
  dfb[colname] <- genrevalue
}
return(dfb)
}

trainset3<- process(df3,trainset3)
testset3<- process(df3,testset3)

#################################################################################
#split Director
process<- function(dfa,dfb){
glist <- NULL
totaln <- length(dfa$Director)
for (n in 1:totaln){
    gen <- unlist(strsplit(dfa$Director[n], ", "))
    glist<- (c(glist,gen))
}

glist<-names(sort(summary(as.factor(glist)), decreasing=T)[1:11])
glist<-glist[-1]



# for every genre in glist, create a new colum in dataframe:
for (g in glist){
  colname <- g
  genrevalue<- NULL
  for (i in 1:length(dfb$Director)){
    if (grepl(g, dfb$Director[i], fixed=TRUE))
      genrevalue <- c(genrevalue,1)
    else genrevalue <- c(genrevalue,0)
  }
  dfb[colname] <- genrevalue
}
return(dfb)
}

trainset3<- process(df3,trainset3)
testset3<- process(df3,testset3)

#################################################################################
#split Actors
process<- function(dfa,dfb){
glist <- NULL
totaln <- length(dfa$Actors)
for (n in 1:totaln){
    gen <- unlist(strsplit(dfa$Actors[n], ", "))
    glist<- (c(glist,gen))
}

glist<-names(sort(summary(as.factor(glist)), decreasing=T)[1:11])
glist<-glist[-1]


# for every genre in glist, create a new colum in dataframe:
for (g in glist){
  colname <- g
  genrevalue<- NULL
  for (i in 1:length(dfb$Actors)){
    if (grepl(g, dfb$Actors[i], fixed=TRUE))
      genrevalue <- c(genrevalue,1)
    else genrevalue <- c(genrevalue,0)
  }
  dfb[colname] <- genrevalue
}
return(dfb)
}

trainset3<- process(df3,trainset3)
testset3<- process(df3,testset3)


#################################################################################
names(testset3)<-gsub("\\s","",names(testset3))
names(testset3)<-gsub("-","",names(testset3))
names(trainset3)<-gsub("\\s","",names(trainset3))
names(trainset3)<-gsub("-","",names(trainset3))

#now data is ready for lm()
#create a function to do training on the data
trainmodel3<- function(data){
 model<- lm((paste("Gross", "~", paste(colnames(data)[c(40:71)], collapse = "+"), sep = "" )), data=data )
 #model using all columns, include wins, nominations, top 10 genre, top 10 director and top 10 actor
 return(model)
}
# models with j=20 various % of the data, repeat i=10 times for each %
trealgross<- testset3[37]
trainrmse<- matrix( 1:40,  nrow=20, ncol=2) 
testrmse<- matrix( 1:40,  nrow=20, ncol=2) 
rmse<- 0
for (j in 1:20){ # use 20 different percentages
  lista<-NULL
  listb<-NULL
  for (i in 1:20){ # repeat 20 times
  select<-selectrows(trainset3,j/20)
  data<- trainset3[select,]
  M3<- trainmodel3(data)
  rmse<-sqrt(mean(residuals(M3)^2))
  lista<-c(lista,rmse)
  tpgross<- as.data.frame( predict(M3,testset3))
  trmse<-0
  rss<- 0
  for (a in 1:length(testset3[,1])){
  e<- (tpgross[a,1]-trealgross[a,1])^2
  rss<- rss+e
  }
  trmse<- sqrt(rss/length(testset3[,1]))
  listb<-c(listb,trmse)
  }
  trainrmse[j,2]<-mean(lista) # this records all train rmse
  trainrmse[j,1]<- j*5 # this records the percentage of data used
  testrmse[j,2]<-mean(listb) # this records all train rmse
  testrmse[j,1]<- j*5 # this records the percentage of data used
}
testresult3<- as.data.frame(testrmse)
colnames(testresult3)<- c("percentage","RMSE")
trainresult3<- as.data.frame(trainrmse)
colnames(trainresult3)<- c("percentage","RMSE")

ggplot(testresult3,aes(y=RMSE, x = percentage)) +  geom_point(data=testresult3,aes(y=RMSE, x = percentage, colour='test')) +  geom_line(data=testresult3,aes(y=RMSE, x = percentage, colour='test')) + geom_point(data=trainresult3,aes(y=RMSE, x = percentage, colour='train')) +  geom_line(data=trainresult3,aes(y=RMSE, x = percentage, colour='train'))





```

##  Numeric and categorical variables

Here we try to improve the prediction quality as much as possible by using both numeric and non-numeric variables

```{r}
# Build & evaluate model 4 (numeric & converted non-numeric variables)


trainmodel4<- function(data){
  model<- lm((paste("Gross~Budget+Year+Runtime+I(Runtime^2)+I(Runtime^3)+imdbRating*imdbVotes+tomatoUserRating*tomatoUserReviews+budgetbin+", paste(colnames(data)[c(40:71)], collapse = "+"), sep = "" )), data=data )
 #model using all columns, include wins, nominations, top 10 genre, top 10 director and top 10 actor
 return(model)
}


# models with j=20 various % of the data, repeat i=10 times for each %
trealgross<- testset3[37]
trainrmse<- matrix( 1:40,  nrow=20, ncol=2) 
testrmse<- matrix( 1:40,  nrow=20, ncol=2) 
rmse<- 0
for (j in 1:20){ # use 20 different percentages
  lista<-NULL
  listb<-NULL
  for (i in 1:20){ # repeat 20 times
  select<-selectrows(trainset3,j/20)
  data<- trainset3[select,]
  M3<- trainmodel4(data)
  rmse<-sqrt(mean(residuals(M3)^2))
  lista<-c(lista,rmse)
  tpgross<- as.data.frame( predict(M3,testset3))
  trmse<-0
  rss<- 0
  for (a in 1:length(testset3[,1])){
  e<- (tpgross[a,1]-trealgross[a,1])^2
  rss<- rss+e
  }
  trmse<- sqrt(rss/length(testset3[,1]))
  listb<-c(listb,trmse)
  }
  trainrmse[j,2]<-mean(lista) # this records all train rmse
  trainrmse[j,1]<- j*5 # this records the percentage of data used
  testrmse[j,2]<-mean(listb) # this records all train rmse
  testrmse[j,1]<- j*5 # this records the percentage of data used
}
testresult3<- as.data.frame(testrmse)
colnames(testresult3)<- c("percentage","RMSE")
trainresult3<- as.data.frame(trainrmse)
colnames(trainresult3)<- c("percentage","RMSE")

ggplot(testresult3,aes(y=RMSE, x = percentage)) +  geom_point(data=testresult3,aes(y=RMSE, x = percentage, colour='test')) +  geom_line(data=testresult3,aes(y=RMSE, x = percentage, colour='test')) + geom_point(data=trainresult3,aes(y=RMSE, x = percentage, colour='train')) +  geom_line(data=trainresult3,aes(y=RMSE, x = percentage, colour='train'))



```


## Additional features

Now we try creating additional features such as interactions (e.g. `is_genre_comedy` x `is_budget_greater_than_3M`) or deeper analysis of complex variables (e.g. text analysis of full-text columns like `Plot`).

```{r}
# Build & evaluate model 5 (numeric, non-numeric and additional features)



trainmodel4<- function(data){
  model<- lm((paste("Gross~Budget*imdbRating+ Budget+Year+Runtime+I(Runtime^2)+I(Runtime^3)+imdbRating*imdbVotes+tomatoUserRating*tomatoUserReviews+budgetbin+", paste(colnames(data)[c(40:71)], collapse = "+"), sep = "" )), data=data )
 #model using all columns, include wins, nominations, top 10 genre, top 10 director and top 10 actor
 return(model)
}


# models with j=20 various % of the data, repeat i=10 times for each %
trealgross<- testset3[37]
trainrmse<- matrix( 1:40,  nrow=20, ncol=2) 
testrmse<- matrix( 1:40,  nrow=20, ncol=2) 
rmse<- 0
for (j in 1:20){ # use 20 different percentages
  lista<-NULL
  listb<-NULL
  for (i in 1:20){ # repeat 20 times
  select<-selectrows(trainset3,j/20)
  data<- trainset3[select,]
  M3<- trainmodel4(data)
  rmse<-sqrt(mean(residuals(M3)^2))
  lista<-c(lista,rmse)
  tpgross<- as.data.frame( predict(M3,testset3))
  trmse<-0
  rss<- 0
  for (a in 1:length(testset3[,1])){
  e<- (tpgross[a,1]-trealgross[a,1])^2
  rss<- rss+e
  }
  trmse<- sqrt(rss/length(testset3[,1]))
  listb<-c(listb,trmse)
  }
  trainrmse[j,2]<-mean(lista) # this records all train rmse
  trainrmse[j,1]<- j*5 # this records the percentage of data used
  testrmse[j,2]<-mean(listb) # this records all train rmse
  testrmse[j,1]<- j*5 # this records the percentage of data used
}
testresult3<- as.data.frame(testrmse)
colnames(testresult3)<- c("percentage","RMSE")
trainresult3<- as.data.frame(trainrmse)
colnames(trainresult3)<- c("percentage","RMSE")

ggplot(testresult3,aes(y=RMSE, x = percentage)) +  geom_point(data=testresult3,aes(y=RMSE, x = percentage, colour='test')) +  geom_line(data=testresult3,aes(y=RMSE, x = percentage, colour='test')) + geom_point(data=trainresult3,aes(y=RMSE, x = percentage, colour='train')) +  geom_line(data=trainresult3,aes(y=RMSE, x = percentage, colour='train'))


```

